### 2021-10-15 plos_latex_template.tex
Rocío: edits to intro and conclusions

% Rocío: Main message of the paragraph: Why we are writing this piece
Conferences are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions. 
However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields.
Lack of representation and unwelcoming--or overtly aggressive--environments hinder participation (or future participation) of people who could otherwise become active community members. 
An exclusionary conference experience can divert career paths, affect lives, and drive people out of academia \cite{biggsAcademicConferenceChilly2018}.
Efforts to address systematic discrimination in conferences and make them inclusive spaces can have a meaningful impact on our communities and the lives and careers of their members.

% Rocío: Main message of the paragraph: We do this for communities and people. What do we mean by communities and why are they important? (Why are they important may need some text)
This article suggests rules to pivot traditional conferences towards diversity and inclusion, and strive to build more inclusive and welcoming communities. 
Some scholarly or technological communities are formally structured as learned societies (e.g. the International Geographical Union or the Royal Statistical Society), others are composed by networks of less formal local groups (e.g. R User Groups or Python User Groups), and others may not have any established structure. 
For the latter, conferences may play even a bigger role in shaping the community, since there is no other organization or institution to gather the community members.

% Rocío: Main message of the paragraph: Roadmap of the paper
The rules written here are directed to people who are part of a stable meetings committee that oversees the site/location selection process, or that coordinates with the local organizers of conferences.
The rules can also be helpful to local/virtual organizers who desire to make an inclusive conference starting at the planning stage.
These tips stem from the authors' experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for everyone. 
Here, we share the lessons learned within the past year of organizing this global useR! 2021, summarized as ten simple rules towards an inclusive conference.
We have organized these rules in three groups (Figure \ref{fig:diagram}).
%sara: i think here is a good place to add the diagram: https://docs.google.com/drawings/d/1iS1pLc9OldLMe_jhNT6OGIbT9s5KvlHpUoDd19vHkCY/edit?usp=sharing
%sara: still need to better format figure size and get the alt text right :P
Rules 1, 2 and 10 refer to pillars of an inclusive conference: embracing diversity in all its dimensions, creating a safe and welcoming environment for everyone, and making the conference part of a long-term process for inclusion.
The next two rules are focused on the people that participate in the conference: 
Rule 3 refers to the importance of working with an inclusive and diverse organizing team, and Rule 4 concerns the necessity of removing implicit and systemic bias from spotlight roles like keynote speakers, other presenters, program committee members, or other session chairs. 
Rules 5 to 9 are rules about actions regarding components of the conference that should be carefully planned for: an online component, accessibility to people with disabilities, language inclusiveness, a welcoming communication strategy, and financial resources to support inclusion. 
These rules apply equally to academic, industry, or mixed conferences; the rules are inspired by a global experience but also apply at the regional or local level.
% Rocío: the last sentence was copied from the abstract because we need a closing sentence.
% Rocío: Could add a line introducing Fig. in https://docs.google.com/drawings/d/1iS1pLc9OldLMe_jhNT6OGIbT9s5KvlHpUoDd19vHkCY/edit?usp=sharing and the whole structure [x]

@@

@@

% Rocío: main idea: unbias roles and give spotlight to the roles--and people who have contributed in roles--that are not stereotypically categorized as success
In academia and the industry, not only does the list of people considered in spotlight roles is usually biased, but the spotlight roles themselves are.
Scientific publication and development of software, are often regarded with higher value than community building or teaching, which may be equally if not more challenging, less rewarded, and usually tackled by women, people of color, people with disabilities and from other minoritized groups \cite{cheng2020x+}.
Defy the stereotypical criteria for success by acknowledging these community practices and the people behind them.
You can, for instance, give more space in your conference to talks about community building, and reframe the awards ceremony to acknowledge those who prepared accessible slides and presentations (see Rule \ref{rule_accessibility}), for being mindful of inclusiveness. 
Finally, do not restrict people from marginalized groups or community-builders to talk or work only in issues related to diversity and inclusion; recognize their areas of expertise and interest.

@@

% Rocío: main idea: Networking may seem the weakest part of online conference, but it doesn't have to be. 
% Rocío: I'm still doubting if this should be the 2nd paragraph, and I think that it needs some editing.
Networking and socializing have been mentioned as challenging aspects of online conferences, mostly because we are used to interactions at in-person social settings such as coffee or lunch breaks \cite{salibaGettingGripsOnline2020, roosOnlineConferencesNew2020}. 
However, in-person settings might not be comfortable or appealing to everyone for socializing; 
some of these spaces may be exclusionary (e.g. galas or dinner nights at expensive venues), and common place for code of conduct violations to occur \cite{auroraHowRespondCode2019}. 
On the other hand, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012}).
Organizers of online conferences should invest time in creating opportunities to meet and bond virtually, respecting people's limits, preferences, and remembering that `the usual' does not necessarily work for everyone, and that one single networking activity will never serve the whole community. 
It is worth trying new and varied activities that might work with subgroups of attendees.
Some ideas in this line are: offering the option of written chat only, instead of voice or video conversations, opening events with teamwork like trivia, offer some events that can be enjoyed passively like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session.

@@

% Rocío: Could start with an introduction (or put somewhere else) that online can easily make conferences more inclusive for many people, but not necessarily for people with disabilities, and that they can be easily left behind unless we actively work on their inclusion. 
% Liz: let me know if you want me to take a shot at this.adding reference marks2021meeting

Conferences are among the least accessible spaces that people with disabilities may encounter \cite{priceAccessImaginedConstruction2009}. Even when conferences implement other inclusive practices, the participation of people with disabilities is often overlooked \cite{marks2021meeting}. This is one key aspect where having someone in the organizing team allows them to take part in the decisions from the beginning (see \textbf{Rule \ref{rule_organizing_team}}), as thinking about disability or simulating it is not a substitute for real-life experience \cite{costanzachockDesign2020}. Accessibility requires time and early decision-making \cite{irishIncreasingParticipationUsing2020} and should not be dealt with at the last minute. Conversely, inaccessible decisions are hard to correct, e.g., when finding out too late that a venue is inaccessible for people who use wheelchairs.

Importantly, accessibility practices are inclusive not only for people with disabilities but can be beneficial to a broad spectrum of people. For instance, having captions is helpful to deaf and hard-of-hearing people, non-native speakers, and everyone in general.

If the conference is in person, the venue must be accessible for people who use wheelchairs, have important signs in Braille, and a sound system compatible with hearing devices, just to name a few important features. Presenters should always speak into a microphone. Take care proactively of invisible disabilities, for example, by providing quiet spaces for privacy and noise-free conversations, menus that include sugar-free, gluten-free options.

Regardless of the conference format --online, in person, or hybrid-- all images used in the communication strategy of the conference should have alternative text, including the website. Likewise, any videos should have both captions and sound.

Platforms for conference registration and abstract submission, websites, and chat platforms --if used-- should be screen-reader friendly and keyboard accessible, with low technology requirements (hardware, software, and internet connection). 
Captioning or interpretation for presentations should be available in more than one language if possible. 
Do not forget about making your social events accessible and making sure that the materials presented are accessible for all. This includes physical activities that should not restrict participation based on body type or ability.

These aspects should be tested well in advance of going live.

%teach about accessibility
Some practices, such as making accessible slides and presentations, are not yet common practice and will require great efforts from the presenters if they are not used to them. 
For that reason, the organizing team should provide accessibility guidelines for slides and presentations, encourage their use, and be available for any questions they may have. Among other things, the guidelines should ask for raw and accessible material to the talks before the conference, in screen-reader-frendly formats (e.g., R Markdown, HTML, or \TeX{} files). Accessibility in slides includes alternative text, and written code instead of print screen.

If presentations are pre-recorded, the speakers should include their video and ensure that their face is visible so that deaf and hard-of-hearing people can read their lips if needed (see \url{https://user2021.r-project.org/participation/accessibility/} for example).

Most importantly, accessibility practices are not afterthoughts that can be dealt with at the last minute. 
They require time and early decision-making \cite{irishIncreasingParticipationUsing2020}, (see Rule \ref{rule_financial} about allocating resources). Conversely, inaccessible decisions are hard to course-correct, e.g., when finding out too late that a venue is inaccessible for people who use wheelchairs.

@@

The ten rules stated here can be adapted depending on the conference format and settings.
When organizing useR! 2021, we engaged in most of the practices mentioned here, and learned others along the way, so we share them here as part of our learning process. 
We organized useR! during a global pandemic, and as a team, this was a challenging journey. 
Very early on, we tried to recognize that our resources were limited (both in time of the organizing team and financially). 
We had to let go of certain aspects that were important to many of us. 
From that experience, this is our last message: your conference may not become perfectly inclusive and accessible, but the changes you make will make a difference.
Also, keep working to improve inclusion in the next edition of the conference! to document and balance your work, design surveys, ask for feedback from attendees and organizers. Transfer your constructed knowledge to the next organizing team, and adapt year-to-year to include more dimensions of diversity, to address technical and social changes, and to include insights from the new organizing team. Always remember inclusion is an ambition that organizers can and should continually pursue by allowing room for learning and improvement.
If more conferences and domains apply these rules, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead.
And you will make a change towards healthier, stronger, and more inclusive communities.

% Try not to burn you and the team out trying to make it perfect.

>>>>

% Rocío: Main message of the paragraph: Why we are writing this piece
Conferences, from the Latin \textit{conferre} or `to bring together', are spaces to meet and reconnect with members from a specific community, learn about advances in the field, and share our recent contributions. 
However, conferences are likely to reproduce the systematic discrimination occurring in other spaces in our fields. 
Lack of representation and unwelcoming--or overtly aggressive--environments exclude people from the opportunities for learning, producing knowledge, and becoming active community members that conferences can provide to the non-marginalized \cite{hendersonThoughtfulGatheringsGendering2020}.
Exclusionary conference experiences can even divert career paths, affect lives, and drive people out of academia \cite{biggsAcademicConferenceChilly2018, hendersonThoughtfulGatheringsGendering2020}. 
This message is not new. 
Barriers such as unattainable registration costs, sexism, and ableism, among others, have already been exposed and discussed in the literature \cite{arendDisparityConferenceRegistration2019, biggsAcademicConferenceChilly2018, depickerRethinkingInclusionDisability2020a, irishIncreasingParticipationUsing2020}, 
and some proposals for more inclusive conferences have been put into practice \cite{gichoraTenSimpleRules2010a, levitisCenteringInclusivityDesign2021, atkinsonJournalMedicine20202021, foramittiVirtuesVirtualConferences2021, ninerBetterWhomLeveling2021, rabyMovingAcademicConferences2021}, with a primary focus an the online format to open doors for inclusion.

% Rocío: Main message of the paragraph: Roadmap of the paper
This article puts the focus on practices for inclusive conferences, whether they are virtual or in-person, though we advocate for an online component. 
The rules written here are directed to people who are part of a stable meetings committee that oversees the site/location selection process, or that coordinates with the local organizers of conferences.
The rules can also be helpful to local/virtual organizers who desire to make an inclusive conference starting at the planning stage.
These tips stem from the authors' experience of organizing useR! 2021, a virtual and global statistical computing conference for users and developers of the R programming language \cite{r_core_team_2021}. 
We embraced the challenge of organizing a high-quality virtual conference in the context of the COVID-19 pandemic and making it a kind, inclusive, and accessible experience for everyone. 
Here, we share the lessons learned within the past year of organizing useR! 2021, summarized as ten simple rules towards an inclusive conference.
The rules are organized in three groups (Figure \ref{fig:diagram}). %original source: https://docs.google.com/drawings/d/1iS1pLc9OldLMe_jhNT6OGIbT9s5KvlHpUoDd19vHkCY/edit?usp=sharing
%sara: still need to better format figure size and get the alt text right :P
Rules 1, 2 and 10 refer to pillars of an inclusive conference: embracing diversity in all its dimensions, creating a safe and welcoming environment for everyone, and making the conference part of a long-term process for inclusion.
The next two rules are focused on the people that participate in the conference: 
Rule 3 refers to the importance of working with an inclusive and diverse organizing team, and Rule 4 concerns the necessity of removing implicit and systemic bias from spotlight roles like keynote speakers, other presenters, program committee members, or other session chairs. 
Rules 5 to 9 are rules about components of the conference that should be carefully planned for: an online component, accessibility to people with disabilities, language inclusiveness, a welcoming communication strategy, and financial resources to support inclusion. 
These rules apply equally to academic, industry, or mixed conferences; the rules are inspired by a global experience but also apply at the regional or local level.

@@

% Rocío: I'm leaving this out now and seeing if we don't need this description of communities
%% Rocío: Main message of the paragraph: We do this for communities and people. What do we mean by communities and why are they important? (Why are they important may need some text)
%This article suggests rules to pivot traditional conferences towards diversity and inclusion, and strive to build more inclusive and welcoming communities. 
%Some scholarly or technological communities are formally structured as learned societies (e.g. the International Geographical Union or the Royal Statistical Society), others are composed by networks of less formal local groups (e.g. R User Groups or Python User Groups), and others may not have any established structure. 
%For the latter, conferences may play even a bigger role in shaping the community, since there is no other organization or institution to gather the community members.

@@

% Rocío: main idea: unbias roles and give spotlight to the roles--and people who have contributed in roles--that are not stereotypically categorized as success
In academia and the industry, not only does the list of people considered in spotlight roles is usually biased, but the spotlight roles themselves are.
Scientific publication and development of software, are often regarded with higher value than community building or teaching, which may be equally if not more challenging, less rewarded, and usually tackled by women, people of color, people with disabilities and from other minoritized groups \cite{cheng2020x+, burfordHomelinessMeantHaving2020}.
Defy the stereotypical criteria for success by acknowledging these community practices and the people behind them.
You can, for instance, give more space in your conference to talks about community building as an area of knowledge production, and reframe the awards ceremony to acknowledge those who prepared accessible slides and presentations (see Rule \ref{rule_accessibility}), for being mindful of inclusiveness. 
Finally, do not restrict people from marginalized groups or community-builders to talk or work only on issues related to diversity and inclusion; recognize their areas of expertise and interest.

@@

% Rocío: main idea: Networking may seem the weakest part of online conference, but it doesn't have to be. 
% Rocío: I'm still doubting if this should be the 2nd paragraph, and I think that it needs some editing.
Networking and socializing have been mentioned as challenging aspects of online conferences, mostly because we are used to interactions at in-person social settings such as coffee or lunch breaks \cite{salibaGettingGripsOnline2020, roosOnlineConferencesNew2020}. 
However, in-person settings might not be comfortable or appealing to everyone for socializing; 
some of these spaces may be exclusionary (e.g. galas or dinner nights at expensive venues), and common place for code of conduct violations to occur \cite{auroraHowRespondCode2019}. 
On the other hand, there is evidence that virtual communication before face-to-face interaction can make people from minoritized groups feel more included, thus participate more (e.g. \cite{trianaDoesOrderFacetoFace2012, blackEngenderingBelongingThoughtful2020}). 
Organizers of online conferences should invest time in creating opportunities to meet and bond virtually, respecting people's limits, preferences, and remembering that `the usual' does not necessarily work for everyone, and that one single networking activity will never serve the whole community. 
It is worth trying new and varied activities that might work with subgroups of attendees.
Some ideas in this line are: offering the option of written chat only, instead of voice or video conversations, opening events with teamwork like trivia, offer some events that can be enjoyed passively like movies, yoga sessions, or art displays, where attendants can choose to just sit and enjoy without talking, or have a chat channel to comment on their experiences during the session.

@@

Conferences are among the least accessible spaces that people with disabilities may encounter \cite{priceAccessImaginedConstruction2009}. Even when conferences implement other inclusive practices, the participation of people with disabilities is often overlooked \cite{marks2021meeting}. This is one key aspect where having someone in the organizing team allows them to take part in the decisions from the beginning (see \textbf{Rule \ref{rule_organizing_team}}), as thinking about disability or simulating it are not substitutes for real-life experience \cite{costanzachockDesign2020}. Planning for accessibility requires time and early decision-making \cite{irishIncreasingParticipationUsing2020} and should not be dealt with at the last minute. Key decisions in this respect are hard to correct, e.g., when finding out too late that a venue is inaccessible for people who use wheelchairs.

For in-person conferences, the venue should comply with common accessibility standards, such as being adequate for people who use wheelchairs, have signs in Braille, and a sound system compatible with hearing devices and live interpretation, just to name a few important features. In addition to this, the organizers should take care proactively of invisible disabilities, for example, by providing quiet spaces for privacy and noise-free conversations, provide chairs in open spaces, and menus that include sugar-free, gluten-free options.

Regardless of the conference format, all platforms (website, chat, conference administration tools) and images used for the communication strategy of the conference should be screen-reader friendly and keyboard accessible, and have alternative text. Any videos should have captions and a transcript. The organizing team should provide accessibility guidelines for slides and presentations, encourage their use, and be available for any questions presenters and attendees may have. Slide decks should be made available beforehand, either in webpages or available for download to ensure that everybody can follow their content during the presentations.

Accessibility practices must include social events and networking, and include activities that do not restrict participation based on body type or ability. Importantly, all these accessibility practices are inclusive not only for people with disabilities but to a broader spectrum of people. For instance, captions are helpful for non-native speakers, having the material available for download helps attendeed with low bandwidth connection, etc. Accessibility efforts should be explicitly displayed in the official communications of the conference (see \textbf{Rule \ref{rule_communication}}).

@@

This article suggests rules to pivot traditional conferences towards diversity and inclusion, and strive to build more inclusive and welcoming communities. 
The ten rules stated here can be adapted depending on the conference format and settings.
We organized useR! during a global pandemic, and as a team, this was a challenging journey. 
We engaged in most of the practices mentioned here and learned others along the way. 
Organizing a conference and implementing inclusive practices are both learning experiences;
do not expect to achieve everything, set priorities and remember rule 10: diversity and inclusion are processes. 
If you are part of a stable meetings committee, you could encourage organizers to follow these rules. 
If more conferences and domains apply them, the process will get more streamlined, straightforward, and mainstream to adapt with minimal overhead.
And you will make a change towards healthier, stronger, and more inclusive communities.
